Note on: Connections between Smoothing Splines and Laplacian smoothing

### 2021-05-21 notes/misc/splines_graph_laplacians_connections.tex
Updated notes on connections between smoothing splines and graph Laplacian smoothing.

Suppose now that in addition to the \emph{labeled} data $(x_1,y_1),\ldots,(x_n,y_n)$, we observe $m$ \emph{unlabeled} design points $x_{n + 1},\ldots,x_{n + m}$ Let $N = n + m$, let $G_{N}$ denote the weighted graph over $x_1,\ldots,x_N$, and $L_{N,\varepsilon} \in \Reals^{N \times N}$ denote the graph Laplacian of $G_{N}$. Thus $L_{N,\varepsilon}$ satisfies,for any $u \in \Reals^N$,
\begin{equation}
\label{eqn:graph_laplacian_all}
\bigl(L_{N,\varepsilon}u\bigr)_i := \frac{1}{N \varepsilon^{d + 2}}\sum_{j = 1}^{N} (u_i - u_j) \eta\biggl(\frac{\|x_i - x_j\|}{\varepsilon}\biggr).
\end{equation}
The first-order graph Laplacian smoothing estimator $\wh{f}_n \in \Reals^n$ defined in \textcolor{red}{(?)} can be naturally generalized to an estimator $\wh{f}_N$ in $\Reals^N$ (that is, a solution over both labeled and unlabeled points), according to
\begin{equation}
\label{eqn:ssl_graph_laplacian_smoothing_0}
\wh{f}_N := \argmin_{f \in \Reals^N} \|Y - f\|_n^2 + \lambda \dotp{L_{N,\varepsilon}f}{f}_N.
\end{equation}

@@

Finally, let $G_n^{(u)} := ([n], W_n^{(u)})$ be a weighted graph, with the $(i,j)$th entry of weighted edge matrix $W_n^{(u)}$ given by
\begin{equation*}
\bigl(W_{n}^{(u)}\bigr)_{ij} := \sum_{k,\ell = m + 1}^{m + n} \eta\biggl(\frac{\|x_i - x_k\|}{\varepsilon}\biggr) \cdot \eta\biggl(\frac{\|x_j - x_{\ell}\|}{\varepsilon}\biggr) \cdot \Bigl(m\varepsilon^{d + 2}L_{m,\varepsilon} + D_{m,n}\Bigr)_{k,\ell}^{-1}.
\end{equation*}
Let $L_{n,\varepsilon}^{(u)}$ be the corresponding Laplacian, meaning
\begin{equation*}
\bigl(L_{n,\varepsilon}^{(u)}u\bigr)_{i} := \frac{1}{m\varepsilon^{d + 2}} \sum_{j = 1}^{n} (u_i - u_j) W_{n,ij}^{(u)}.
\end{equation*}
The proof of Proposition~\ref{prop:ssl_graph_laplacian_smoothing} is contained in Section~\ref{subsec:pf_ssl_graph_laplacian_smoothing}. 
\begin{proposition}
	\label{prop:ssl_graph_laplacian_smoothing}
	Let $\wh{f}_N$ be the solution to~\eqref{eqn:ssl_graph_laplacian_smoothing_0}. Writing $\wh{f}_N = (\wh{f}_N^{(\ell)},\wh{f}_N^{(u)})$ for $\wh{f}_N^{(\ell)} \in \Reals^n$ and $\wh{f}_N^{(u)} \in \Reals^m$, we have the following.
	\begin{itemize}
		\item The in-sample component $\wh{f}_N^{(\ell)}$ of the first-order Laplacian smoothing estimator $\wh{f}_N$ is given by
		\begin{equation*}
		\wh{f}_N^{(\ell)} := \biggl(I_n + \frac{\rho n}{N}\biggl[\frac{n}{N}L_{n,\varepsilon} + \frac{m}{N}L_{n,\varepsilon}^{(u)}\biggr]\biggr)^{-1} Y.
		\end{equation*}
		\item The out-of-sample component $\wh{f}_N^{(u)}$ of the first-order Laplacian smoothing estimator $\wh{f}_N$ is given by
		\begin{equation*}
		\wh{f}_N^{(u)} := \bigl(D_{m,n} + m\varepsilon^{d + 2}L_{m,\varepsilon}\bigr)^{-1} K_{m,n} \wh{f}_N^{(\ell)}.
		\end{equation*}
	\end{itemize}
\end{proposition}
Proposition~\ref{prop:ssl_graph_laplacian_smoothing} shows two things. First, it shows that at the labeled points $x_1,\ldots,x_n$, the estimator $\wh{f}_N$ is exactly an in-sample graph Laplacian smoothing estimator, but with the crucial distinction that the graph Laplacian is the (weighted) average of $L_{n,\varepsilon}$ and $L_{n,\varepsilon}^{(u)}$. The second thing Proposition~\ref{prop:ssl_graph_laplacian_smoothing} reveals is that at unlabeled points, the estimator $\wh{f}_N$ (or rather, its restriction to unlabeled points $\wh{f}_N^{(u)}$) satisfies the graph Poisson equation
\begin{equation*}
(D_{m,n} + m\varepsilon^{d + 2}L_{m,\varepsilon})\wh{f}_N^{(u)} = K_{m,n} \wh{f}_N^{(\ell)}.
\end{equation*}
In words, the out-of-sample component $\wh{f}_N^{(u)}$ is the solution to an inverse problem, in which the right hand side is a \emph{smoothening} of the in-sample component $\wh{f}_N^{(\ell)}$. This smoothening is a crucial aspect of the out-of-sample component. The role it plays is most clearly understood when there is exactly one unlabeled point.

\subsection{One unlabeled point}
Suppose $m = 1$. Then Proposition~\ref{prop:ssl_graph_laplacian_smoothing} simplifies. In particular, the out-of-sample component of $\wh{f}_N$ can be written in terms of the Nadaraya-Watson kernel smoother $T_{n,\varepsilon}$ defined by $T_{n,\varepsilon}u(x) := 1/d_{n,\varepsilon}(x) K_{(x)} u :=  1/d_{n,\varepsilon}(x) \cdot \sum_{i = 1}^{n} u_i \eta(\|x_i - x\|/\varepsilon)$. Here we have written $K_{(x)}$ for the column vector of kernel evaluations at labeled points, $K_{(x)} := (\eta(\|x_1 - x\|/\varepsilon),\ldots,\eta(\|x_n - x\|/\varepsilon))$. 
\begin{corollary}
	\label{cor:ssl_graph_laplacian_smoothing_one_unlabeled}
	Let $\wh{f}_N$ be the solution to~\eqref{eqn:ssl_graph_laplacian_smoothing_0}. Writing $\wh{f}_N = (\wh{f}_N^{(\ell)},\wh{a})$ for $\wh{f}_N^{(\ell)} \in \Reals^n$ and $\wh{a} \in \Reals$, we have the following.
	\begin{itemize}
		\item The in-sample component $\wh{f}_N^{(\ell)}$ of the first-order Laplacian smoothing estimator $\wh{f}_N$ is given by
		\begin{equation}
		\label{eqn:ssl_graph_laplacian_smoothing_one_unlabeled_1}
		\wh{f}_N^{(\ell)} := \biggl(I_n + \frac{\rho n}{n + 1}\biggl[\frac{n}{n + 1}L_{n,\varepsilon} + \frac{1}{n + 1}L_{n,\varepsilon}^{(1)}\biggr]\biggr)^{-1} Y.
		\end{equation}
		\item The out-of-sample component $\wh{a}$ of the first-order Laplacian smoothing estimator $\wh{f}_N$ is given by
		\begin{equation}
		\label{eqn:ssl_graph_laplacian_smoothing_one_unlabeled_2}
		\wh{a} = \bigl(T_{n,\varepsilon}\wh{f}_N^{(\ell)}\bigr)(x_{n + 1}) = \frac{1}{d_{n,\varepsilon}(x_{n + 1})} K_{(x_{n + 1})}\wh{f}_N^{(\ell)}.
		\end{equation}
	\end{itemize}
\end{corollary}
From~\eqref{eqn:ssl_graph_laplacian_smoothing_one_unlabeled_1} we see that, roughly speaking, as long as $L_{n,\varepsilon}^{(u)}$ has a reasonable spectrum, the in-sample component $\wh{f}_N^{(\ell)}$ should be quite close to the Laplacian smoothing estimator $\wh{f}_n$ formed only using the in-sample points. From~\eqref{eqn:ssl_graph_laplacian_smoothing_one_unlabeled_2} we see that the out-of-sample component $\wh{a}$ is constructed by passing a Nadaraya-Watson kernel smoother (with kernel function equal to $\eta$ and bandwidth equal to $\varepsilon$) over $\wh{f}_N^{(\ell)}$.

@@

\subsection{Proof of Theorem~{(?)}}
\label{subsec:pf_ssl_graph_laplacian_smoothing_2}

\paragraph{Strategy.}
The decomposition in Proposition~\ref{prop:ssl_graph_laplacian_smoothing} reveals something of the structure of $\wh{f}_N$. It also suggests a route to analyzing the out-of-sample error $(\wh{a} - f_0(x_{n + 1}))^2$. Concretely, consider the following expansion
\begin{align*}
\wh{a} - f_0(x_{n + 1}) & = \wh{a} - \frac{1}{d_{n,\varepsilon}(x_{n + 1})}K_{(x_{n + 1})} \Ebb_n\wh{f}_N^{(\ell)} + \\
& \quad \frac{1}{d_{n,\varepsilon}(x_{n + 1})}K_{(x_{n + 1})} \Ebb_n\wh{f}_N^{(\ell)} - \frac{1}{d_{n,\varepsilon}(x_{n + 1})}K_{1,n} f_0 + \\
& \quad \frac{1}{d_{n,\varepsilon}(x_{n + 1})}K_{(x_{n + 1})} f_0 - f_0(x_{n + 1}) \\
\Longrightarrow \Ebb_n\bigl[|\wh{a} - f_0(x_{n + 1})|^2\bigr]& \leq \underbrace{\|T_{n,\varepsilon}(\wh{f}_N^{(\ell)} - \mathbb{E}_{n}\wh{f}_N^{(\ell)})\|_P^2}_{\textrm{variance}} + \underbrace{\|T_{n,\varepsilon}(\mathbb{E}_{n}\wh{f}_N^{(\ell)} - f_0)\|_P^2}_{\textrm{bias}} + \underbrace{\|T_{n,\varepsilon}f_0 - f_0\|_P^2}_{\textrm{noiseless kernel smoothing}}
\end{align*}
Here $\Ebb_n$ denotes the expectation conditional on the labeled design points $x_1,\ldots,x_n$. Now we proceed to bound each of these terms, with high probability.

\paragraph{Term 1: Variance.}
Let $\wt{L}_{n,\varepsilon} = n/(n + 1) L_{n,\varepsilon} + 1/(n + 1) L_{n,\varepsilon}^{(u)}$. Denoting the smoother matrix $S_n := [I_n + \rho n/(n + 1)\wt{L}_{n,\varepsilon}]^{-1}$, we can write the variance term as
\begin{equation*}
\|T_{n,\varepsilon}(\wh{f}_N^{(\ell)} - \mathbb{E}_{n}\wh{f}_N^{(\ell)})\|_P^2 =  \int_{\Omega} \Bigl(T_{n,\varepsilon}(S_nw)(x)\Bigr)^2 \,dP(x).
\end{equation*}

Note that for any $i,j$ in $1,\ldots,n$, we have
\begin{equation*}
\Ebb_n\bigl[(S_nw)_i \cdot (S_nw)_j\bigr] = (S_n^2)_{ij}.
\end{equation*}
Therefore, at a given $x \in \Omega$ it follows that
\begin{equation*}
\Ebb_n\Bigl[\bigl(T_{n,\varepsilon}S_nw(x)\bigr)^2\Bigr] = \frac{1}{d_{n,\varepsilon}(x)^2} K_{(x)}^{\top} S_n^2 K_{(x)} \leq \frac{1}{d_{n,\varepsilon}(x)^2} \|S_n\|_{\textrm{op}}^2 \cdot \|K_{(x)}\|_2^2 \leq \frac{1}{d_{n,\varepsilon}(x)^2} \cdot \|K_{(x)}\|_2^2.
\end{equation*}
Now by Lemma~\ref{lem:max_degree_deviation}, we have that for both $K(x) = \eta(x)$ and $K = \bigl(\eta(x)\bigr)^2$, with probability at least $1 - C_2\exp\bigl(-c_2 \delta^2 \cdot n \varepsilon^d\bigr)$,
\begin{equation}
\label{pf:out_of_sample_laplacian_regularization_1}
\sup_{x \in \Omega} \biggl|\frac{d_{n,\varepsilon}(x; K)}{n} - d_{P,\varepsilon}(x; K)\biggr| \leq \delta \cdot \inf_{x \in \Omega} d_{P,\varepsilon}(x; K),
\end{equation}
and thus
\begin{equation*}
\sup_{x \in \Omega}\frac{1}{d_{n,\varepsilon}(x)^2} \cdot \|K_{(x)}\|_2^2 \leq \biggl[\frac{(1 + \delta)}{(1 - \delta)}\biggr]^2 \sup_{x \in \Omega} \frac{1}{nd_{P,\varepsilon}(x)^2} \int \bigl[\eta(\|x'-x\|/\varepsilon)\bigr]^2 \,dP(x') \leq \biggl[\frac{(1 + \delta)}{(1 - \delta)}\biggr]^2 \frac{\|\eta\|_{\infty}}{n p_{\min} \varepsilon^d}.
\end{equation*}
Integrating with respect to $P$, and exchanging expectation with integration thus gives that with probability at least $1 - C_2\exp\bigl(-c_2 \delta^2 \cdot n \varepsilon^d\bigr)$,
\begin{equation*}
\|T_{n,\varepsilon}(\wh{f}_N^{(\ell)} - \mathbb{E}_{n}\wh{f}_N^{(\ell)})\|_P^2 \leq \biggl[\frac{(1 + \delta)}{(1 - \delta)}\biggr]^2 \frac{\|\eta\|_{\infty} p_{\max}}{n p_{\min} \varepsilon^d}.
\end{equation*}

\paragraph{Term 2: Bias.}
\textcolor{red}{(TODO)}.

\paragraph{Term 3: Noiseless kernel smoothing.}
This is a standard analysis. \textcolor{red}{From our paper on Laplacian eigenmaps (equation (107) and Lemma~19)}, we have that
\begin{equation*}
\|T_{n,\varepsilon}f_0 - f_0\|_P^2 \leq \frac{C}{\delta} \cdot \biggl(\frac{\varepsilon^2}{n\varepsilon^d} + \varepsilon^2\biggr) |f|_{H^1(\Omega)}^2 \leq \frac{C}{\delta} \cdot \varepsilon^2 |f|_{H^1(\Omega)}^2
\end{equation*}
with probability at least $1 - C\exp(-cn\varepsilon^d)$.

>>>>

Suppose now that in addition to the \emph{labeled} data $(x_1,y_1),\ldots,(x_n,y_n)$, we observe $m$ \emph{unlabeled} design points $x_{n + 1},\ldots,x_{n + m}$. Let $N = n + m$, let $G_{N}$ denote the weighted graph over $x_1,\ldots,x_N$, and $L_{N,\varepsilon} \in \Reals^{N \times N}$ denote the graph Laplacian of $G_{N}$. Thus $L_{N,\varepsilon}$ satisfies,for any $u \in \Reals^N$,
\begin{equation}
\label{eqn:graph_laplacian_all}
\bigl(L_{N,\varepsilon}u\bigr)_i := \frac{1}{N \varepsilon^{d + 2}}\sum_{j = 1}^{N} (u_i - u_j) \eta\biggl(\frac{\|x_i - x_j\|}{\varepsilon}\biggr).
\end{equation}
The first-order graph Laplacian smoothing estimator $\wh{f}_n \in \Reals^n$ defined in \textcolor{red}{(?)} can be naturally generalized to an estimator $\wh{f}_N$ in $\Reals^N$ (that is, a solution over both labeled and unlabeled points), according to
\begin{equation}
\label{eqn:ssl_graph_laplacian_smoothing_0}
\wh{f}_N := \argmin_{f \in \Reals^N} \|Y - f\|_n^2 + \lambda \dotp{L_{N,\varepsilon}f}{f}_N.
\end{equation}

@@

Finally, let $G_n^{(u)} := ([n], W_n^{(u)})$ be a weighted graph, with the $(i,j)$th entry of weighted edge matrix $W_n^{(u)}$ given by
\begin{equation}
\label{eqn:induced_insample_graph}
\bigl(W_{n}^{(u)}\bigr)_{ij} := \sum_{k,\ell = m + 1}^{m + n} \eta\biggl(\frac{\|x_i - x_k\|}{\varepsilon}\biggr) \cdot \eta\biggl(\frac{\|x_j - x_{\ell}\|}{\varepsilon}\biggr) \cdot \Bigl(m\varepsilon^{d + 2}L_{m,\varepsilon} + D_{m,n}\Bigr)_{k,\ell}^{-1}.
\end{equation}
Let $L_{n,\varepsilon}^{(u)}$ be the corresponding Laplacian, meaning
\begin{equation*}
\bigl(L_{n,\varepsilon}^{(u)}u\bigr)_{i} := \frac{1}{m\varepsilon^{d + 2}} \sum_{j = 1}^{n} (u_i - u_j) W_{n,ij}^{(u)}.
\end{equation*}
The proof of Proposition~\ref{prop:ssl_graph_laplacian_smoothing} is contained in Section~\ref{subsec:pf_ssl_graph_laplacian_smoothing}. 
\begin{proposition}
	\label{prop:ssl_graph_laplacian_smoothing}
	Let $\wh{f}_N$ be the solution to~\eqref{eqn:ssl_graph_laplacian_smoothing_0}. Writing $\wh{f}_N = (\wh{f}_N^{(\ell)},\wh{f}_N^{(u)})$ for $\wh{f}_N^{(\ell)} \in \Reals^n$ and $\wh{f}_N^{(u)} \in \Reals^m$, we have the following.
	\begin{itemize}
		\item The in-sample component $\wh{f}_N^{(\ell)}$ of the first-order Laplacian smoothing estimator $\wh{f}_N$ is given by
		\begin{equation*}
		\wh{f}_N^{(\ell)} := \biggl(I_n + \frac{\rho n}{N}\biggl[\frac{n}{N}L_{n,\varepsilon} + \frac{m}{N}L_{n,\varepsilon}^{(u)}\biggr]\biggr)^{-1} Y.
		\end{equation*}
		\item The out-of-sample component $\wh{f}_N^{(u)}$ of the first-order Laplacian smoothing estimator $\wh{f}_N$ is given by
		\begin{equation*}
		\wh{f}_N^{(u)} := \bigl(D_{m,n} + m\varepsilon^{d + 2}L_{m,\varepsilon}\bigr)^{-1} K_{m,n} \wh{f}_N^{(\ell)}.
		\end{equation*}
	\end{itemize}
\end{proposition}
Proposition~\ref{prop:ssl_graph_laplacian_smoothing} shows two things. First, it shows that at the labeled points $x_1,\ldots,x_n$, the estimator $\wh{f}_N$ is exactly an in-sample graph Laplacian smoothing estimator, but with the crucial distinction that the graph Laplacian is the (weighted) average of $L_{n,\varepsilon}$ and $L_{n,\varepsilon}^{(u)}$. The second thing Proposition~\ref{prop:ssl_graph_laplacian_smoothing} reveals is that at unlabeled points, the estimator $\wh{f}_N$ (or rather, its restriction to unlabeled points $\wh{f}_N^{(u)}$) satisfies the graph Poisson equation
\begin{equation*}
(D_{m,n} + m\varepsilon^{d + 2}L_{m,\varepsilon})\wh{f}_N^{(u)} = K_{m,n} \wh{f}_N^{(\ell)}.
\end{equation*}
In words, the out-of-sample component $\wh{f}_N^{(u)}$ is the solution to an inverse problem, in which the right hand side is a \emph{smoothening} of the in-sample component $\wh{f}_N^{(\ell)}$. This smoothening is a crucial aspect of the out-of-sample component. The role it plays is most clearly understood when there is exactly one unlabeled point.

\subsection{One unlabeled point}
Suppose $m = 1$. Then Proposition~\ref{prop:ssl_graph_laplacian_smoothing} simplifies. In particular, the out-of-sample component of $\wh{f}_N$ can be written in terms of the Nadaraya-Watson kernel smoother $T_{n,\varepsilon}$ defined by $T_{n,\varepsilon}u(x) := 1/d_{n,\varepsilon}(x) K_{(x)} u :=  1/d_{n,\varepsilon}(x) \cdot \sum_{i = 1}^{n} u_i \eta(\|x_i - x\|/\varepsilon)$. Here we have written $K_{(x)}$ for the column vector of kernel evaluations at labeled points, $K_{(x)} := (\eta(\|x_1 - x\|/\varepsilon),\ldots,\eta(\|x_n - x\|/\varepsilon))$. 
\begin{corollary}
	\label{cor:ssl_graph_laplacian_smoothing_one_unlabeled}
	Let $\wh{f}_N$ be the solution to~\eqref{eqn:ssl_graph_laplacian_smoothing_0}. Writing $\wh{f}_N = (\wh{f}_N^{(\ell)},\wh{a})$ for $\wh{f}_N^{(\ell)} \in \Reals^n$ and $\wh{a} \in \Reals$, we have the following.
	\begin{itemize}
		\item The in-sample component $\wh{f}_N^{(\ell)}$ of the first-order Laplacian smoothing estimator $\wh{f}_N$ is given by
		\begin{equation}
		\label{eqn:ssl_graph_laplacian_smoothing_one_unlabeled_1}
		\wh{f}_N^{(\ell)} := \biggl(I_n + \frac{\rho n}{n + 1}\biggl[\frac{n}{n + 1}L_{n,\varepsilon} + \frac{1}{n + 1}L_{n,\varepsilon}^{(u)}\biggr]\biggr)^{-1} Y.
		\end{equation}
		\item The out-of-sample component $\wh{a}$ of the first-order Laplacian smoothing estimator $\wh{f}_N$ is given by
		\begin{equation}
		\label{eqn:ssl_graph_laplacian_smoothing_one_unlabeled_2}
		\wh{a} = \bigl(T_{n,\varepsilon}\wh{f}_N^{(\ell)}\bigr)(x_{n + 1}) = \frac{1}{d_{n,\varepsilon}(x_{n + 1})} K_{(x_{n + 1})}\wh{f}_N^{(\ell)}.
		\end{equation}
	\end{itemize}
\end{corollary}
From~\eqref{eqn:ssl_graph_laplacian_smoothing_one_unlabeled_1} we see that, roughly speaking, as long as $L_{n,\varepsilon}^{(u)}$ has a reasonable spectrum, the in-sample component $\wh{f}_N^{(\ell)}$ should be quite close to the Laplacian smoothing estimator $\wh{f}_n$ formed only using the in-sample points. From~\eqref{eqn:ssl_graph_laplacian_smoothing_one_unlabeled_2} we see that the out-of-sample component $\wh{a}$ is constructed by passing a Nadaraya-Watson kernel smoother (with kernel function equal to $\eta$ and bandwidth equal to $\varepsilon$) over $\wh{f}_N^{(\ell)}$.

Interestingly, the kernel smoothing step in~\eqref{eqn:ssl_graph_laplacian_smoothing_one_unlabeled_2} allows us to make dramatically stronger claims about the out-of-sample squared error $|\wh{a} - f_0(x_{n + 1})|^2$ than we could about the in-sample mean squared error $\|\wh{f}_n - f_0\|_n^2$. We will assume the following.
\begin{assumption}[Design distribution $P$.]
	\label{asmp:design2}
	The design distribution $P$ is supported on an open, bounded set $\Omega \subseteq \Rd$, which has $C^{1}$ boundary $\partial \Omega$. The distribution $P$ admits a $C^{1}(\Omega)$ density $p$ with respect to Lebesgue measure $\nu$, which is bounded away from $0$ and $\infty$, meaning that for some $p_{\min} < 1 < p_{\max}$
	\begin{equation*}
	0 \leq p_{\min} \leq p(x) \leq p_{\max} < \infty,\quad\textrm{for all $x \in \Omega$.}
	\end{equation*}
\end{assumption}  
\begin{theorem}
	\label{thm:out_of_sample_laplacian_regularization}
	Suppose $(x_1,y_1),\ldots,(x_n,y_n),x_{n + 1}$ are sampled from a distribution $P$ which satisfies Assumption~\ref{asmp:design2}, that the regression function $f_0$ is \textcolor{red}{$M$-Lipschitz} on $\Omega$. If the estimator $\wh{f}_N$ in~\eqref{eqn:ssl_graph_laplacian_smoothing_0} is computed with $0 < \rho \leq M^2(M^2n)^{-2/(2 + d)}$ and $\varepsilon = n^{-1/(2 + d)}$, then 
	\begin{equation*}
	\bigl|\wh{a} - f_0(x_{n + 1})\bigr|^2 \leq C\frac{M^2}{\delta^2}(M^2n)^{-2/(2 + d)},
	\end{equation*}
	with probability at least $1 - C_1\delta - C_2\exp(-c_2n\varepsilon^d)$.
\end{theorem}

@@

\subsection{Proof of Theorem~\ref{thm:out_of_sample_laplacian_regularization}}
\label{subsec:pf_ssl_graph_laplacian_smoothing_2}

\paragraph{Strategy.}
The decomposition in Proposition~\ref{prop:ssl_graph_laplacian_smoothing} reveals something of the structure of $\wh{f}_N$. It also suggests a route to analyzing the out-of-sample error $(\wh{a} - f_0(x_{n + 1}))^2$. Let $S_N$ and $S_n$ denote the smoother matrices such that $\wh{f}_N^{(\ell)} = S_NY$ and $\wh{f}_n = S_nY$; explicitly
\begin{equation*}
S_N := \Biggl(I_n + \frac{\rho n}{N}\biggl[\frac{n}{N}L_{n,\varepsilon} + \frac{1}{N}L_{n,\varepsilon}^{(u)}\biggr]\Biggr)^{-1}, \quad \textrm{and} \quad S_n := \Biggl(I_n + \frac{\rho n}{N}L_{n,\varepsilon}\Biggr)^{-1}.
\end{equation*}
Now we consider the following decomposition
\begin{align*}
\wh{a} - f_0(x_{n + 1}) & = T_{n,\varepsilon}S_NY(x_{n + 1}) - T_{n,\varepsilon}S_Nf_0(x_{n + 1}) + T_{n,\varepsilon}S_Nf_0(x_{n + 1}) - T_{n,\varepsilon}S_nf_0(x_{n + 1}) + T_{n,\varepsilon}S_nf_0(x_{n + 1}) - T_{n,\varepsilon}f_0(x_{n + 1}) + \\
& \quad T_{n,\varepsilon}f_0(x_{n + 1}) - f_0(x_{n + 1}) \\
& =  \underbrace{T_{n,\varepsilon}S_Nw(x_{n + 1})}_{\textrm{Term 1}} + \underbrace{T_{n,\varepsilon}[(S_N - S_n)f_0](x_{n + 1})}_{\textrm{Term 2}} +  \underbrace{T_{n,\varepsilon}S_nf_0(x_{n + 1}) - T_{n,\varepsilon}f_0(x_{n + 1})}_{\textrm{Term 3}} + \underbrace{T_{n,\varepsilon}f_0(x_{n + 1}) - f_0(x_{n + 1})}_{\textrm{Term 4}}.
\end{align*}
The first of these terms is the contribution of the noise in $Y$. The second of these terms is the perturbation caused by the different smoother matrices. The third of these terms is the bias due to Laplacian regularization. The fourth term is the error inherent to kernel smoothing of noiseless samples. Now we proceed to upper bound the error induced by each of these terms.

\paragraph{Term 1: Noise contribution.}
Let $\Ebb_N$ denote the expectation conditional on the design points $x_1,\ldots,x_{n + 1}$. For any $i,j$ in $1,\ldots,n$, we have
\begin{equation*}
\Ebb_N\bigl[(S_Nw)_i \cdot (S_nw)_j\bigr] = (S_N^2)_{ij}.
\end{equation*}
Also, $\|S_N\|_{\textrm{op}} \leq 1$. 
Therefore
\begin{equation}
\label{pf:out_of_sample_laplacian_regularization_0}
\Ebb_N\Bigl[\bigl(T_{n,\varepsilon}S_Nw(x_{n + 1})\bigr)^2\Bigr] = \frac{1}{d_{n,\varepsilon}(x_{n + 1})^2} K_{(x_{n + 1})}^{\top} S_N^2 K_{(x_{n + 1})} \leq \frac{1}{d_{n,\varepsilon}(x_{n + 1})^2} \|S_N\|_{\textrm{op}}^2 \cdot \|K_{(x_{n + 1})}\|_2^2 \leq \frac{1}{d_{n,\varepsilon}(x_{n + 1})^2} \cdot \|K_{(x_{n + 1})}\|_2^2.
\end{equation}
Now by Lemma~\ref{lem:max_degree_deviation}, we have that for both $K(x) = \eta(x)$ and $K = \bigl(\eta(x)\bigr)^2$, with probability at least $1 - C_2\exp\bigl(-c_2 \delta^2 \cdot n \varepsilon^d\bigr)$,
\begin{equation}
\label{pf:out_of_sample_laplacian_regularization_1}
\sup_{x \in \Omega} \biggl|\frac{d_{n,\varepsilon}(x; K)}{n} - d_{P,\varepsilon}(x; K)\biggr| \leq \delta \cdot \inf_{x \in \Omega} d_{P,\varepsilon}(x; K),
\end{equation}
and thus
\begin{equation*}
\sup_{x \in \Omega}\frac{1}{d_{n,\varepsilon}(x)^2} \cdot \|K_{(x)}\|_2^2 \leq \biggl[\frac{(1 + \delta)}{(1 - \delta)}\biggr]^2 \sup_{x \in \Omega} \frac{1}{nd_{P,\varepsilon}(x)^2} \int \bigl[\eta(\|x'-x\|/\varepsilon)\bigr]^2 \,dP(x') \leq \biggl[\frac{(1 + \delta)}{(1 - \delta)}\biggr]^2 \frac{\|\eta\|_{\infty}}{n p_{\min} \varepsilon^d}.
\end{equation*}
Plugging this back in to~\eqref{pf:out_of_sample_laplacian_regularization_0} gives
\begin{equation*}
\Ebb_N\Bigl[\bigl(T_{n,\varepsilon}S_Nw(x_{n + 1})\bigr)^2\Bigr] \leq \biggl[\frac{(1 + \delta)}{(1 - \delta)}\biggr]^2 \frac{\|\eta\|_{\infty}}{n p_{\min} \varepsilon^d}.
\end{equation*}
Thus by Markov's inequality, with probability at least $1 - \beta - C_2\exp\bigl(-c_2 \delta^2 \cdot n \varepsilon^d\bigr)$
\begin{equation*}
\bigl(T_{n,\varepsilon}S_Nw(x_{n + 1})\bigr)^2 \leq \frac{1}{\beta}\biggl[\frac{(1 + \delta)}{(1 - \delta)}\biggr]^2 \frac{\|\eta\|_{\infty}}{n p_{\min} \varepsilon^d} = C_1\frac{1}{n\varepsilon^d}.
\end{equation*}

\paragraph{Term 2: Perturbation.}
From the Cauchy-Schwarz inequality, we have that for any $u \in \Reals^n$,
\begin{equation*}
|T_{n,\varepsilon}u(x_{n + 1})|^2 = \biggl|\frac{1}{d_{n,\varepsilon}(x_{n + 1})} \sum_{i = 1}^{n} u_i \eta\biggl(\frac{\|x_i - x_{n + 1}\|}{\varepsilon}\biggr)\biggr|^2 \leq \frac{1}{d_{n,\varepsilon}(x_{n + 1})} \sum_{i = 1}^{n} |u_i|^2 \eta\biggl(\frac{\|x_{i} - x_{n + 1}\|}{\varepsilon}\biggr) \leq \|u\|_{\infty}^2.
\end{equation*}
Thus to upper bound Term 2 we focus our attention on upper bounding $\|(S_N - S_n)f_0\|_{\infty}$.

Since $S_N$ and $S_n$ are symmetric, and thus $S_NS_n = S_nS_N$, it follows that
\begin{equation*}
(S_N - S_n)f_0 = (S_NS_nS_n^{-1} - S_nS_NS_N^{-1})f_0 = S_N S_n(S_n^{-1} - S_N^{-1})f_0.
\end{equation*}
Consequently we can upper bound the $\infty$-norm $\|(S_N - S_n)f_0\|_{\infty}$ by
\begin{equation*}
\|(S_N - S_n)f_0\|_{\infty} \leq \|S_N\|_{\infty} \|S_n\|_{\infty} \|(S_n^{-1} - S_N^{-1})f_0\|_{\infty}.
\end{equation*}
Here we recall that the $\infty$-operator norm of a matrix is $\|A\|_{\infty} = \max\{\|Au\|_{\infty}:\|u\|_{\infty} = 1\}$.

We first upper bound $\|(S_n^{-1} - S_N^{-1})f_0\|_{\infty}$, using the decomposition
\begin{equation*}
S_n^{-1} - S_N^{-1} = \rho \biggl[\biggl(1 - \frac{n^2}{N^2}\biggr)L_{n,\varepsilon} - \frac{n}{N^2}L_{n,\varepsilon}^{(u)}\biggr],
\end{equation*}
and therefore
\begin{equation}
\label{pf:out_of_sample_laplacian_regularization_2}
\|(S_n^{-1} - S_N^{-1})f_0\|_{\infty} \leq \frac{\rho}{n} \Bigl(\|L_{n,\varepsilon}f_0\|_{\infty} + \|L_{n,\varepsilon}^{(u)}f_0\|_{\infty} \Bigr).
\end{equation}
From H\"{o}lder's inequality and the Lipschitz property of $f_0$, we deduce both that
\begin{align*}
\|L_{n,\varepsilon}f_0\|_{\infty} & = \max_{i = 1,\ldots,n} \biggl| \frac{1}{n\varepsilon^{d + 2}}\sum_{j = 1}^{n} \bigl(f_0(x_i) - f_0(x_j)\bigr) \eta\biggl(\frac{\|x_i - x_j\|}{\varepsilon}\biggr) \biggr| \\
& \leq \max_{i = 1,\ldots,n}  \frac{M}{n\varepsilon^{d + 1}}\sum_{j = 1}^{n} \eta\biggl(\frac{\|x_i - x_j\|}{\varepsilon}\biggr) \\
& \leq M\frac{1}{\varepsilon^{d + 1}},
\end{align*}
and
\begin{align*}
\|L_{n,\varepsilon}^{(u)}f_0\|_{\infty} & = \max_{i = 1,\ldots,n} \biggl| \frac{1}{\varepsilon^{d + 2}} \sum_{j = 1}^{n} \bigl(f_0(x_i) - f_0(x_j)\bigr) W_{n,ij}^{(u)}\biggr| \\
& \leq \max_{i = 1,\ldots,n} \frac{M}{\varepsilon^{d + 1}} \sum_{j = 1}^{n} |W_{n,ij}^{(u)}|\\
& \leq \frac{M}{\varepsilon^{d + 1}} \|\eta\|_{\infty},
\end{align*}
where we recall the definition of $W_{n}^{(u)}$ from~\eqref{eqn:induced_insample_graph}. Plugging these upper bounds back into~\eqref{pf:out_of_sample_laplacian_regularization_2} gives
\begin{equation*}
\|(S_n^{-1} - S_N^{-1})f_0\|_{\infty} \leq 2 \|\eta\|_{\infty}\frac{\rho}{n\varepsilon^{d + 1}}M.
\end{equation*}
On the other hand, it is not hard to see that $S_N$ and $S_n$ are contractions in $L^{\infty}(\Reals^n)$, meaning that $\|S_N\|_{\infty},\|S_n\|_{\infty} \leq 1$. We establish that $S_n$ is a contraction, and the exact same reasoning will hold with respect to $S_N$ as well. For any $u \in \Reals^n$, assuming without loss of generality that $u_1 = \|u\|_{\infty}$, we have that
\begin{equation*}
(S_n^{-1}u)_1 = u_1 + (\rho L_{n,\varepsilon}u)_1 \geq u_1,
\end{equation*}
which implies that $\|S_n^{-1}u\|_{\infty} \geq \|u\|_{\infty}$. Thus for any $u \in \Reals^n$, we also have that
\begin{equation*}
\|u\|_{\infty} = \|S_n^{-1}S_nu\|_{\infty} \geq \|S_nu\|_{\infty},
\end{equation*}
demonstrating that $S_n$ is a contraction.
